---
title: "Activity: Graph Replication"
author: "Matthew Beckman"
date: ""
output: 
  html_document:
    fig_height: 3
    fig_width: 5
---

```{r}
# libraries
library(tidyverse)
library(DataComputing)
library(mosaicData)
```


# Grading

Assignment is worth a total of 10 points:

- [3 points] Problem 5.3 graph replication  
- [3 points] Problem 5.4 graph replication   
- Exploration of `diamonds` data  
    - [1 point] `ggplot2` expression that successfully produces a graphical display of the `diamonds` data.  
    - [1 point] graphic must display at least 3 variables from the `diamonds` data.  
    - [2 points] written narrative describing something learned/observed about the `diamonds` data using your graphic.  

# Activity

Reproduce plots for Problem 5.3 & Problem 5.4 in the Data Computing Text.


###  Problem 5.3

The "yes" and "no" in the gray bars refer to whether or not the person is pregnant.


```{r}
# Note: I've added some spacing to make it a little easier to read the function
 ggplot(data = NCHS, aes(x = bmi)) + 
  geom_density(adjust = 0.4) + 
  aes(colour = sex) + 
  facet_wrap( ~ pregnant, ncol = 4) + 
  labs(title = "") + 
  theme(legend.position = "right") 
```




###  Problem 5.4
```{r}
# Note: I've added some spacing to make it a little easier to read the function
ggplot(data = CPS85, aes(x = exper, y = wage)) + 
  geom_point() + 
  aes(colour = married) + 
  scale_x_log10() + 
  scale_y_log10() + 
  facet_wrap( ~ sector, ncol = 4)
```



### Your Turn: explore the `diamonds` data

*Task: Produce an interesting graphic of the `diamonds` data set.  The graphic should represent AT LEAST 3 variables.  Write a few sentences to describe what you have learned.*

```{r}
data("diamonds")

```




### Optional Extension: programming style

The helper functions are great, but they produce ugly code.  Clean up the `ggplot2` code that reproduces the required plots so that it aligns with the R Programming Style Guide Appendix in the *DataComputing* eBook <https://dtkaplan.github.io/DataComputingEbook/>.  


