---
title: "HELPmiss (Data Verbs) Activity"
author: "Matthew Beckman"
output: html_notebook
---


## Set Up:

```{r message=FALSE}
rm(list = ls())     # clean up your R environment

# load packages
library(tidyverse)  # includes lots of data verbs like `group_by()` and `summarise()`
library(mosaicData) # includes the `HELPmiss` data set 


# Load the `HELPmiss` data set into our RStudio environment
data("HELPmiss", package = "mosaicData")

```


## Task 1: Data Description

*Write several sentences (or a bullet list) describing the HELP Study and the resulting `HELPmiss` data.  Your description should investigate basic data provenance (e.g., Who, What, When, Where, Why, How), explain the setting for the data, specify what each case represents in the data, and remark on inclusion/exclusion criteria.* 


- investigate by searching R help documentation
- there's a research paper cited where additional detail is provided



## Task 2: Basic Summaries 

*Produce one or more R expressions involving `summarize()` and `HELPmiss` to address each of the following prompts.*

- number of people (cases) in `HELPmiss` study
- total number of times in the past 6 months entered a detox program (measured at baseline) for all the people in `HELPmiss` (silly)
- mean time (in days) to first use of any substance post-detox for all the people in `HELPmiss`

```{r}
HELPmiss %>%
  summarise( cases = n() )

HELPmiss %>%
  summarise( totalDetoxBaseline = sum(e2b, na.rm = TRUE)) 

HELPmiss %>%
  summarise( avgDaysRelapse = mean(daysanysub, na.rm = TRUE))
```


## Task 3: Group Summaries  

*Repeat task 2 above, but add code chunks to calculate result group-by-group according to each prompt.  Be sure to show all R code and write a sentence or two about what you observe in the results.* 


- males versus females
- homeless or not
- substance
- break down the homeless versus housed further, by sex
- break down the homeless versus housed further, by substance

### males versus females
```{r}
HELPmiss %>%
  group_by(sex) %>%
  summarise( cases = n(),
             totalDetoxBaseline = sum(e2b, na.rm = TRUE),
             avgDaysRelapse = mean(daysanysub, na.rm = TRUE))
```

### homeless or not
```{r}
HELPmiss %>%
  group_by(homeless) %>%
  summarise( cases = n(),
             totalDetoxBaseline = sum(e2b, na.rm = TRUE),
             avgDaysRelapse = mean(daysanysub, na.rm = TRUE))
```


### substance
```{r}
HELPmiss %>%
  group_by(substance) %>%
  summarise( cases = n(),
             totalDetoxBaseline = sum(e2b, na.rm = TRUE),
             avgDaysRelapse = mean(daysanysub, na.rm = TRUE))
```

### homeless versus housed broken down by sex
```{r}
HELPmiss %>%
  group_by(homeless, sex) %>%
  summarise( cases = n(),
             totalDetoxBaseline = sum(e2b, na.rm = TRUE),
             avgDaysRelapse = mean(daysanysub, na.rm = TRUE))
```

### homeless versus housed broken down by substance
```{r}
HELPmiss %>%
  group_by(homeless, substance) %>%
  summarise( cases = n(),
             totalDetoxBaseline = sum(e2b, na.rm = TRUE),
             avgDaysRelapse = mean(daysanysub, na.rm = TRUE))
```



## Task 4: Data Visualization & Observations

*Include one or more interesting plots of the data involving at least 3 variables per plot. Write a few sentences to explain the story that your plot tells about these data.  You can expand on the relationships that you studied in Task 2, or you can explore a different group of variables in `HELPmiss` that show something interesting.*  



* Plots vary
* Almost any plot that you want to post here would be fine as long as it involves at least 3 variables.



