---
title: "Activity: Data Verbs HELPrct"
author: "Matthew Beckman"
date: ""
output: html_notebook 
---


# Grading

- Assignment is worth a total of 10 points
- [3 points] Task 1: 
    - Data description
    - `summarise()` tasks
- [5 points] Task 2: 
    - one point per `group_by()` item 
- [2 points] Task 3: 
    - one point for including a useful plot with at least 3 variables represented
    - one point for including comments


# Set Up

```{r}
# load packages
library(tidyverse)

# The HELPrct data are available in the mosaicData package
library(mosaicData)

# Load the HELPrct data set into our RStudio environment
data("HELPrct")

# Use View(HELPrct) in the console to open a tab in RStudio and look at the raw data directly
```



# Task 1: 

### Data Description

[write a couple of sentences describing the `HELPrct` data and the study that collected them. (hint: use the "Help" search bar in RStudio)]

### Basic Summaries
`summarise()` : Find an expression involving `summarize()` and `HELPrct` that will produce the following.

- number of people (cases) in `HELPrct` study
- total number of times in the past 6 months entered a detox program (measured at baseline) for all the people in `HELPrct` (silly)
- mean time (in days) to first use of any substance post-detox for all the people in `HELPrct`

```{r}
HELPrct %>%
  summarise( cases = n() )

HELPrct %>%
  summarise( totalDetoxBaseline = sum(e2b, na.rm = TRUE)) 

HELPrct %>%
  summarise( avgDaysRelapse = mean(daysanysub, na.rm = TRUE))
```



# Task 2:
`group_by()` : repeat task 1 above, but calculating the results group-by-group for:

- males versus females
- homeless or not
- substance
- break down the homeless versus housed further, by sex
- break down the homeless versus housed further, by substance

### males versus females
```{r}
HELPrct %>%
  group_by(sex) %>%
  summarise( cases = n(),
             totalDetoxBaseline = sum(e2b, na.rm = TRUE),
             avgDaysRelapse = mean(daysanysub, na.rm = TRUE))
```

### homeless or not
```{r}
HELPrct %>%
  group_by(homeless) %>%
  summarise( cases = n(),
             totalDetoxBaseline = sum(e2b, na.rm = TRUE),
             avgDaysRelapse = mean(daysanysub, na.rm = TRUE))
```


### substance
```{r}
HELPrct %>%
  group_by(substance) %>%
  summarise( cases = n(),
             totalDetoxBaseline = sum(e2b, na.rm = TRUE),
             avgDaysRelapse = mean(daysanysub, na.rm = TRUE))
```

### homeless versus housed broken down by sex
```{r}
HELPrct %>%
  group_by(homeless, sex) %>%
  summarise( cases = n(),
             totalDetoxBaseline = sum(e2b, na.rm = TRUE),
             avgDaysRelapse = mean(daysanysub, na.rm = TRUE))
```

### homeless versus housed broken down by substance
```{r}
HELPrct %>%
  group_by(homeless, substance) %>%
  summarise( cases = n(),
             totalDetoxBaseline = sum(e2b, na.rm = TRUE),
             avgDaysRelapse = mean(daysanysub, na.rm = TRUE))
```

# Task 3:

Include one or more interesting plots of the data involving at least 3 variables per plot. Write a few sentances to explain the story that your plot tells about these data.  You can use one of the relationships that you studied in Task 2, or you can explore a different group of variables in the HELPrct that show something interesting.  

* Plots vary
* Almost any plot that you want to post here would be fine as long as it involves at least 3 variables.